---
title: "Deconvolution Algorithm Comparison"
output:
   html_document:
      toc: true
      toc_float:
         collapsed: true
         smooth_scroll: true
params:
   tempdir: ""
   celltype.col: "cell_type"
---
```{r setup2, include=FALSE}
options(width = 1000)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
temp.dir <- params$tempdir
plot.dir <- paste(temp.dir, "report_plots", sep = '/')
celltype.order <- params$celltype.order
celltype.order.sim <- params$celltype.order.sim

benchmark.name <- strsplit(temp.dir, "/")[[1]][
  length(strsplit(temp.dir, "/")[[1]])
]
# read input and processed data
input_raw <- read_data(paste(temp.dir, "/input_data/raw.h5", sep = ""))
sc.counts <- input_raw$sc.counts
sc.pheno <- input_raw$sc.pheno
real.counts <- input_raw$bulk.counts
real.props <- input_raw$bulk.props
validation_set <- read_data(
  paste(temp.dir, "/input_data/validation_set.h5", sep = "")
)
training_set <- read_data(
  paste(temp.dir, "/input_data/training_set.h5", sep = "")
)
training.exprs <- training_set$sc.counts
training.pheno <- training_set$sc.pheno
test.exprs <- validation_set$sc.counts
test.pheno <- validation_set$sc.pheno
sim.bulks <- list(
  bulks = validation_set$real.counts,
  props = validation_set$real.props
)
# load input parameters
input_params <- read_misc_input(
  paste(temp.dir, "input_data/params.h5", sep = "/")
)
genesets <- input_params$genesets
algorithm.names <- input_params$algorithms
function.call <- input_params$function.call
grouping <- input_params$grouping
```

# Deconvolution Benchmark

## Real bulks

### Score plot
```{r score_plot, echo = FALSE,  out.width='100%', out.height = 400}
plotfile <- paste0(plot.dir, "/score_plot_real.pdf")
if(file.exists(plotfile)){
  knitr::include_graphics(plotfile)
}
```

<button  type="button"
   onclick="if(document.getElementById('score_plot_info') .style.display=='none')
              {document.getElementById('score_plot_info') .style.display=''}
            else{document.getElementById('score_plot_info') .style.display='none'}">
  Show/hide description
</button>
<div id="score_plot_info" style="display:none" >

Deconvolution result for real bulks per algorithm and cell type.  
Celltype order (left to right): `r celltype.order`  

The plotted value is the pearson correlation coefficient, coded by rectangle size and color 
(small to big, red to green with increasing performance).   
Negative and undefined scores were set to 0. 
The first column contains the average performance of each algorithm 
across all cell types.
In case of multiple repetitions (default), the individual scores for each cell
type are averaged over all repetitions.
The score specified for each algorithm corresponds to the 'overall' column. 
Uncertainty (standard deviation) of the overall performance in case multiple runs.
</div>

### Scatter plots of quantities
```{r scatter_plots, echo = FALSE, out.width='100%', out.height = 400}
plotfile <- paste0(plot.dir, "/scatter_plots_real.pdf")
if(file.exists(plotfile)){
  knitr::include_graphics(plotfile)
}
```
<button  type="button"
   onclick="if(document.getElementById('real_scatter_plot_info') .style.display=='none')
              {document.getElementById('real_scatter_plot_info') .style.display=''}
            else{document.getElementById('real_scatter_plot_info') .style.display='none'}">
  Show/hide description
</button>
<div id="real_scatter_plot_info" style="display:none" >

Scatterplot of the real cell type quantities (x-axis) and the 
estimated quantities (y-axis) per cell type for each algorithm. Dashed line
marks the identity.

</div>

### Runtime plot
```{r runtime_plot, echo = FALSE, out.width='100%', out.height = 400}
plotfile <- paste0(plot.dir, "/runtime_plot_real.pdf")
if(file.exists(plotfile)){
  knitr::include_graphics(plotfile)
}
```

<button  type="button"
   onclick="if(document.getElementById('real_runtime_info') .style.display=='none')
              {document.getElementById('real_runtime_info') .style.display=''}
            else{document.getElementById('real_runtime_info') .style.display='none'}">
  Show/hide description
</button>
<div id="real_runtime_info" style="display:none" >

Runtime per algorithm. 
Due to possible large differences in runtime the scale on the x-axis is 
transformed to logarithmic scale (log10). 
In case of multiple repetitions, the plotted times are averages over all repetitions.

</div>

### Additional Plots
<button  type="button"
   onclick="if(document.getElementById('additional_info') .style.display=='none')
              {document.getElementById('additional_info') .style.display=''}
            else{document.getElementById('additional_info') .style.display='none'}">
  Show/hide
</button>
<div id="additional_info" style="display:none" >
### Condition number plots
```{r condition_plots, out.width='100%', out.height = 400}
plotfile <- paste0(plot.dir, "/condition_number_plots_real.pdf")
if(file.exists(plotfile)){
  knitr::include_graphics(plotfile)
}
```

Condition number plots:  

1) Average condition numbers of the reference matrices 
used by different algorithms (models).

2) Average score vs condition number of the model  

3) Standard deviation of the score of the model across multiple repetitions 
vs the variation of the model's condition number across repetitions

### Bootstrapped Bulks
```{r bootstrap_plots, out.width='100%', out.height = 400}
plotfile <- paste0(plot.dir, "/bootstrap_plot_real.pdf")
if(file.exists(plotfile)){
  knitr::include_graphics(plotfile)
}
```

Bootstrapping performed to test the robustness of the chosen correlation. 
After deconvolution, a number of bulks is repeatedly chosen at random 
(with replacement) and the score is calculated each time. This gives an
estimate of how robust the deconvolution result is. A large variance indicates 
that the correlation may be driven by outliers.

</div>

## Simulations

### Bulk simulations
### Score plot
```{r bulk_sim_scores, out.width='100%', out.height = 400}
plotfile <- paste0(plot.dir, "/score_plot_simulated.pdf")
if(file.exists(plotfile)){
  knitr::include_graphics(plotfile)
}
```

<button  type="button"
   onclick="if(document.getElementById('sim_score_plot_info') .style.display=='none')
              {document.getElementById('sim_score_plot_info') .style.display=''}
            else{document.getElementById('sim_score_plot_info') .style.display='none'}">
  Show/hide description
</button>
<div id="sim_score_plot_info" style="display:none" >

Deconvolution result for simulated bulks per algorithm and cell type.  
Celltype order (left to right): `r celltype.order.sim`  
The value is the pearson correlation coefficient, coded by rectangle size and color 
(small to big, red to green with increasing performance). 
Negative and undefined scores were set to 0. 
The **overall** column contains the average performance of each algorithm 
across all cell types.
In case of multiple repetitions, the plotted scores are averages over all repetitions. 
The score specified for each algorithm corresponds to the 'overall' column. 
Uncertainty (standard deviation) of the overall performance is specified 
for each algorithm as well.
Bulks were simulated by taking the sum over a fixed number of 
randomly sampled single-cell profiles. The distribution of cell types in the
artificial bulks does not necessarily match the distribution of cell types in the single 
cell data.

</div>

### Scatter plot of quantities
```{r bulk_sim_scatter, out.width='100%', out.height = 400}
plotfile <- paste0(plot.dir, "/scatter_plots_simulated.pdf")
if(file.exists(plotfile)){
  knitr::include_graphics(plotfile)
}
```

<button  type="button"
   onclick="if(document.getElementById('sim_scatter_plot_info') .style.display=='none')
              {document.getElementById('sim_scatter_plot_info') .style.display=''}
            else{document.getElementById('sim_scatter_plot_info') .style.display='none'}">
  Show/hide description
</button>
<div id="sim_scatter_plot_info" style="display:none" >

Comparison of the real cell type quantities in the simulated bulks (x-axis) 
and the estimated quantities per cell type (y-axis) for each algorithm.
Dashed line marks the identity.

</div>

\pagebreak
### Training set size simulation
### Score boxplots
```{r sample_sim, out.width='100%', out.height = 400}
plotfile <- paste0(plot.dir, "/sample_plots_simulated.pdf")
if(file.exists(plotfile)){
  knitr::include_graphics(plotfile)
}
```

<button  type="button"
   onclick="if(document.getElementById('sim_training_plot_info') .style.display=='none')
              {document.getElementById('sim_training_plot_info') .style.display=''}
            else{document.getElementById('sim_training_plot_info') .style.display='none'}">
  Show/hide description
</button>
<div id="sim_training_plot_info" style="display:none" >

Variability of performance of algorithms using different amounts 
of randomly selected training profiles for learning and reference profile creation. 
Overall denotes the average score of all cell types. 
Bulks used for deconvolution are the same as in the simulated score plot above.

</div>

### Geneset simulation

### Runtime lineplot
```{r geneset_sim_time, out.width='100%', out.height = 400}
plotfile <- paste0(plot.dir, "/gene_time_plot_simulated.pdf")
if(file.exists(plotfile)){
  knitr::include_graphics(plotfile)
}
```


<button  type="button"
   onclick="if(document.getElementById('sim_runtime_plot_info') .style.display=='none')
              {document.getElementById('sim_runtime_plot_info') .style.display=''}
            else{document.getElementById('sim_runtime_plot_info') .style.display='none'}">
  Show/hide description
</button>
<div id="sim_runtime_plot_info" style="display:none" >

Runtime of algorithms depending on the amount of genes supplied for deconvolution. 
Each algorithm can perform a separate feature selection on the given gene set 
to obtain the features used in the model.  
Gene sets are ordered by the size of the intersect of features 
available in the gene set and the single-cell data.

</div>

### Score lineplots
```{r geneset_sim, out.width='100%', out.height = 400}
plotfile <- paste0(plot.dir, "/gene_score_plots_simulated.pdf")
if(file.exists(plotfile)){
  knitr::include_graphics(plotfile)
}
```

<button  type="button"
   onclick="if(document.getElementById('sim_lineplot_info') .style.display=='none')
              {document.getElementById('sim_lineplot_info') .style.display=''}
            else{document.getElementById('sim_lineplot_info') .style.display='none'}">
  Show/hide description
</button>
<div id="sim_lineplot_info" style="display:none" >

Performance of algorithms depending on the gene set supplied for deconvolution. 
Each algorithm could perform a separate feature selection on the gene set to 
obtain the features used in the model. 
Gene sets are ordered by the size of the intersect of features available 
in the gene set and the single-cell data. 
Score is the pearson correlation coefficient, averaged over multiple repetitions. 
Errorbars show standard deviation of the score across repetitions.

</div>


### Subtype simulation
### Score plot
```{r subtype_sim, out.width='100%', out.height = 400}
plotfile <- paste0(plot.dir, "/subtype_plot.pdf")
if(file.exists(plotfile)){
  knitr::include_graphics(plotfile)
}
```

<button  type="button"
   onclick="if(document.getElementById('sim_subtype_plot_info') .style.display=='none')
              {document.getElementById('sim_subtype_plot_info') .style.display=''}
            else{document.getElementById('sim_subtype_plot_info') .style.display='none'}">
  Show/hide description
</button>
<div id="sim_subtype_plot_info" style="display:none" >

Average performance of the different algorithms depending on cell type granularity. 
Cell types were simulated from single-cell data using hierarchical clustering 
on a umap embedding of the data. 
The number of distinct cell types increases from left to right. 
Quantities of subtypes of the same cell type obtained by deconvolution are combined 
and their sum is compared to the quantities of the coarse cell types 
(classification that was contained in the single-cell pheno data).

</div>

### Correlation Fits
```{r fits, out.width='100%', out.height = 400}
plotfile <- paste0(plot.dir, "/correlation_fits.pdf")
if(file.exists(plotfile)) {
  knitr::include_graphics(plotfile)
}
plotfile <- paste0(plot.dir, "/fit_offsets.pdf")
if(file.exists(plotfile)) {
  knitr::include_graphics(plotfile)
}
plotfile <- paste0(plot.dir, "/fit_errors.pdf")
if(file.exists(plotfile)) {
  knitr::include_graphics(plotfile)
}
plotfile <- paste0(plot.dir, "/fit_scores.pdf")
if(file.exists(plotfile)) {
  knitr::include_graphics(plotfile)
}
```

<button  type="button"
   onclick="if(document.getElementById('sim_score_plot_info') .style.display=='none')
              {document.getElementById('sim_score_plot_info') .style.display=''}
            else{document.getElementById('sim_score_plot_info') .style.display='none'}">
  Show/hide description
</button>
<div id="sim_score_plot_info" style="display:none" >

Some text for description...
</div>

The following tables contain the numerical values of the averaged fit parameters 
shown in the plots above.

### Offsets
```{r offset_plot, out.width='100%', out.height = 400}
results_file <- paste0(temp.dir, "/results/simulation/scores/deconv_output.h5")
if (file.exists(results_file)) {
  score_results <- rhdf5::h5dump(results_file)
  knitr::kable(score_results$offsets)
}
```

### Error sizes
```{r error_plot, out.width='100%', out.height = 400}
if (exists("score_results")) {
  knitr::kable(score_results$errors)
}
```

## Input data
<button  type="button"
   onclick="if(document.getElementById('input') .style.display=='none')
              {document.getElementById('input') .style.display=''}
            else{document.getElementById('input') .style.display='none'}">
  Show/hide
</button>
<div id="input" style="display:none" >
Parameters of the `benchmark` call:
```{r}
function.call.list <- as.list(function.call)

adjusted.names <- sapply(
  X = names(function.call.list),
  FUN = function(x) {
      if (x != "") {
        return(paste0(x, " ="))
      }else{
        return(x)
      }
    },
  USE.NAMES = FALSE
  )
print(paste(adjusted.names, function.call.list))
```

Single-cell data contained `r nrow(sc.counts)` features for `r ncol(sc.counts)` cells.


Corresponding phenotype data contained labels for `r length(unique(sc.pheno$cell_type))` cell types:
```{r}
  t <- table(sc.pheno[[celltype.col]])
  print(paste(names(t), t, sep = ": "))
```


Used algorithms: `r algorithm.names`


`r ncol(real.counts)` real bulk profiles were available, 
with `r length(intersect(rownames(real.counts), rownames(sc.counts)))` 
overlapping features.  

Ground-truth cell type quantities were available for `r nrow(real.props)` 
cell types.


```{r real_props, echo = FALSE}
	if (exists("real.props") && !is.null(real.props) && length(real.props) > 0) {
    print(t(apply(real.props, 1, summary)))
	}
```
\pagebreak
</div>
